---
title: "Analysis of Variance"
subtitle: "In Class Activity"
author: "Rebecca Hoover"
title-block-banner: true
date: "11/21/2024"
abstract: "This study focuses on statistical methods for mean analysis and hypothesis testing. It covers rejection regions, the Student's t-test for sample comparison, ANOVA for assessing multiple means, and Tukey's test for identifying specific differences among those means."
smooth-scroll: true
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

Metadata:

- Gender (0/1)
- Age (earth years. 🤓)
- Height (cm)
- Pre.weight & weight6weeks (kg)
- Diet (1,2,3)

## Load in Packages and Tidy Data

```{r, message=FALSE, warning=FALSE}
# load the data in here
#Packages
library(tidyverse)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(gridExtra)
#Load data
data <- read.csv("DietData.csv")

data <- data %>%
  mutate(Person_ID = Person,  
         Gender = gender,
         Pre_Weight = pre.weight,
         Weight_6wks = weight6weeks) %>%  
  mutate(Weight_Loss = Pre_Weight - Weight_6wks) %>%
  select(-Person, -gender, -pre.weight, -weight6weeks) %>%
  select(Person_ID, Gender, Age, Height, Diet, Pre_Weight, Weight_6wks, Weight_Loss)

data$Gender <- factor(data$Gender, levels = c(0, 1), labels = c("Female", "Male"))

data$Diet <- factor(data$Diet, levels = c(1, 2, 3), labels = c("Diet 1", "Diet 2", "Diet 3"))


print(summary(data))
```

## Exploratory Data Anlsysis

One of the first things to do is to look at the data and see if there are any obvious things.  Go ahead and explore these data visually.  What do you see?

We have two categorical columns, gender and diet, the rest being of numeric value. We can see if Diet Analysis Goals in weight loss are normally distributed, it is,our categorical data is comparable and our numeric data is normally distributed. We can move forward.

```{r}
# Data visualization
h1 <- ggplot(data, aes(x = Gender, y = Weight_Loss, fill = Gender)) +
  geom_boxplot() + 
  labs(title = "Weight Loss by Gender",
       x = "Gender",
       y = "Weight Loss") +
  scale_fill_manual(values = c("Male" = "cornflowerblue", "Female" = "pink")) +
  theme_minimal() +
  theme(legend.position = "none")

h2 <- ggplot(data, aes(x = Diet, y = Weight_Loss, fill = Diet)) +
  geom_boxplot() +  
  labs(title = "Weight Loss by Diet",
       x = "Diet",
       y = "Weight Loss") +
  scale_fill_brewer(palette = "Set3") +  
  theme_minimal()+
  theme(legend.position = "none")

h3 <- ggplot(data, aes(x = Weight_Loss)) +
  geom_histogram(binwidth = 0.5, color = "blue", fill = "skyblue") + 
  labs(title = "Normal Distribution of Weight Loss",
       x = "Weight Loss (kg)",
       y = "Frequency") +
  theme_minimal()

grid.arrange(h1, h2, h3, nrow=2)
```



## Estimating Mean Values

Table of Weight Loss by Gender and Diet.

```{r, warning=FALSE, message=FALSE}
# Table output
diet_table <- data %>%
  group_by(Gender, Diet) %>%
  summarize(
    Mean_Weight_Loss = mean(Weight_Loss),  
    Median_Weight_Loss = median(Weight_Loss), 
    SD_Weight_Loss = sd(Weight_Loss),         
    Count = n()                               
  )

print(diet_table)
```


## 1-Way Analysis of Variance

*Diet Issues:* 


The underlying linear model.


$$
y_{ij} = \mu + \tau_{Diet, i} + \epsilon_j
$$


Test the null hypothesis, $H_O:$ *There is no effect of diet on weight loss* (e.g., $\tau_{Diet-1} = \tau_{Diet-2} = \tau_{Diet-3} = 0.0$).  Is there evidence for one diet producing more weight loss than the others?  Create an `aov` analysis and assign it to the variable `fit.diet` and examine its contents.

```{r}
# Define model
fit_diet <- aov(Weight_Loss ~ as.factor(Diet), data = data)
summary(fit_diet)
```

Are they all significantly different?  Try the `TukeyHSD()` Interpret the results.

*In comparing the diets, we found that there is no significant difference in weight loss between Diet 1 and Diet 2 (p = 0.9988), indicating that the slight difference observed is not meaningful. However, when comparing Diet 3 to Diet 1, there is a significant difference in weight loss (p = 0.0188), with Diet 3 resulting in more weight loss than Diet 1. Similarly, Diet 3 also shows a significant difference in weight loss compared to Diet 2 (p = 0.0152), again showing that Diet 3 leads to greater weight loss. Overall, the Tukey post-hoc test indicates that Diet 3 results in significantly more weight loss than both Diet 1 and Diet 2, while there is no significant difference between Diet 1 and Diet 2.*

```{r}
# Posthoc test
gobble <- TukeyHSD(fit_diet)
print(gobble)
```

 How much of the variation is explained?  If you notice when you do a summary from a `lm()` (regression) model, it gives you the $R^2$ values directly (remember $R^2 = \frac{SS_{model}}{SS_{Error}}$).  Does `summary()` of your `aov` model give you that?
 
```{r}
# How much variance?
# R-squared for Diet
diet_anova <- tidy(fit_diet)
ss_total <- sum(diet_anova$sumsq)
r2_diet <- diet_anova$sumsq[1] / ss_total
print(paste("R-squared for Diet:", r2_diet))
```

Since I asked the question, the answer is probably no.  Why does it not do this?  Probably for historical reasons, which are a bit of a pain in the backside.  That being said, there are some `tidy` ways to fix this issue.  I'm going to use the `broom` package which allows us to clean up (or tidy if you will) the model objects.  This will take the model object and pull out *all* the ANOVA table stuff and put it into a `tibble`.

```{r,warning=FALSE}
# use your model fit next (I called mine fit.diet)
tidy_diet <- tidy( r2_diet )
tidy_diet
```

Now, since it is all easily accessible, we can calculate the $R^2$ from the new model output.

```{r, warning=FALSE}
# Estimate the variance explained from the raw sums of squares
r2_Diet <- tidy_diet$sumsq[1] / sum( tidy_diet$sumsq )
r2_Diet 
```


 

*Gender:* 

The underlying linear model.


$y_{ij} = \mu + \tau_{gender, i} + \epsilon_j$



Independent of the diet, test the null hypothesis $H_O:$ *There is no difference in weight loss between genders* (e.g., $\tau_{gender-0} = \tau_{gender-2} = 0.0 $).  Is there evidence for one gender being significantly different than another?  How much of the variation is explained (another $R^2$ by gender)?

```{r}
# partition effects
# Define the model: ANOVA for weight loss by gender
fit_gender <- aov(Weight_Loss ~ as.factor(Gender), data = data)

# View the summary of the ANOVA model
summary(fit_gender)

# Tidy the model output for gender
tidy_gender <- tidy(fit_gender)

# Estimate R² for gender model from raw sums of squares
r2_gender <- tidy_gender$sumsq[1] / sum(tidy_gender$sumsq)
r2_gender
```

**How do you interpret these results thus far?**

The p-value of 0.835 indicates that there is no significant difference in weight loss between genders. Additionally, the R-squared value of 0.0005 suggests that gender accounts for only a very small portion (0.05%) of the variation in weight loss, which is essentially negligible.

# Do genders respond differently to diets?


$y_{ijk} = \mu + \tau_{Diet,i} + \tau_{gender,j} + \epsilon_k$

So here we do something a bit different.  We want to simultaneously ask the following questions:

- Do diets influence weight loss?
**The diet has a significant effect on weight loss, as the p-value is less than 0.05.** 
- Do genders influence weight loss?
**Gender does not have a significant effect on weight loss, as the p-value is greater than 0.05.**
- Is there an interaction where different genders respond differently to different diets?
**The interaction between diet and gender is statistically significant, as the p-value is less than 0.05.**

In $R$, this is done as:

```{r}
# diet model and gender model
# Two-way ANOVA model: Diet and Gender interaction
fit_interaction <- aov(Weight_Loss ~ as.factor(Diet) * as.factor(Gender), data = data)

# View the summary of the interaction model
summary(fit_interaction)

# Visualize the interaction between diet and gender
ggplot(data, aes(x = Diet, y = Weight_Loss, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Interaction between Diet and Gender on Weight Loss",
       x = "Diet",
       y = "Weight Loss (kg)") +
  theme_minimal() +
  scale_fill_manual(values = c("pink", "cornflowerblue"))
```


```{r}
# interaction model
# Tidy the model output for interaction
tidy_interaction <- tidy(fit_interaction)

```

What is the $R^2$ for this model?

```{r}
# best model variance explained
# Estimate R² for the interaction model from raw sums of squares
r2_interaction <- tidy_interaction$sumsq[1] / sum(tidy_interaction$sumsq)
r2_interaction
```


# Which Model is Best?

How would you compare the models you generated?  How do you interpret the findings?

*The two-way ANOVA reveals significant effects of diet (p < 0.01) and a diet-gender interaction (p = 0.048), indicating that the effect of diet on weight loss is influenced by gender. The interaction model explains a larger portion of the variance in weight loss compared to diet or gender alone. Specifically, diet has a significant impact on weight loss, with Diet 1 being the most effective. Although gender alone does not significantly affect weight loss, the notable interaction between diet and gender suggests that gender moderates the effect of diet. Overall, the interaction model provides the most comprehensive explanation for the variance in weight loss.*