---
title: "Tidyverse"
subtitle: "In Class Activity"
format: html
execute:
  echo: false
---


## The Question

> Describe the daytime air temperatures at the Rice Rivers Center for February, 2014 by day of the week.


## The Data

For this topic we will use some example data from the [Rice Rivers Center](https://ricerivers.vcu.edu).  These data represent both atmospheric and water data collected from instrumentation on-site.  I have stored these data in a spreadsheet that is shared on Google Drive as a CSV file.

```{r}
library( tidyverse )
url <- "https://docs.google.com/spreadsheets/d/1Mk1YGH9LqjF7drJE-td1G_JkdADOU0eMlrP01WFBT8s/pub?gid=0&single=true&output=csv"
rice <- read_csv( url )
summary( rice )
```

These data represent measurements taken every 15 minutes, 24 hours a day, 7 days a week, 365 days a year.  For brevity, this file contains measurements starting at `r rice$DateTime[1]` and ending at `r rev(rice$DateTime)[1]` (only `r nrow(rice)` records here...).

If you look at the summary of the data above, you will see several things, including:

- Date and time objects are `character`
- Some measurements are in Standard and some in Imperial with units in the same file include both °F and °C, as well as measurements in meters, feet, and inches.  In fact, there are duplication of data columns in different units (guess what kind of correlation they might have...)



```{r}
library( knitr )
library( kableExtra )
library( lubridate)
rice %>%
  select( DateTime, AirTempF) %>%  # select the columns (DateTime, AirTempF)
  mutate( Date = mdy_hms( DateTime, 
                          tz="EST") ) %>%   # mutate to make Date
  select( Date, AirTempF ) %>%
  mutate( Month = month( Date )) %>%
  filter( Month == 2 ) %>%                  # filter for February
  select( -Month ) %>%
  mutate( Hour = hour( Date ) ) %>%      # filter for daytime
  filter( Hour >= 7, Hour <= 18 ) %>% 
  mutate( Weekday = factor( weekdays( Date ), 
                            ordered = TRUE,
                            levels = c("Monday",
                                       "Tuesday",
                                       "Wednesday",
                                       "Thursday",
                                       "Friday",
                                       "Saturday",
                                       "Sunday")) ) %>% # mutate to add Weekday
  mutate( TempC =  (AirTempF - 32) * 5/9    ) %>% 
  group_by( Weekday ) %>%
  summarize( Minimum = min(TempC ),
              Mean = mean( TempC ), 
              Maximum = max( TempC ) )  -> df 

df %>%   # summarize min, max, mean
  kable( digits=3, caption = "My way cool table.") %>%
  kable_classic() 
# kable()


```


Min & Max Turbidity by Month as Kable 

```{r}
rice %>%
  mutate( Date = mdy_hms(DateTime),
          Month = month(Date, abbr = FALSE, label = TRUE) ) %>%
  group_by( Month ) %>%
  summarise( Minimum = min( Turbidity_ntu, na.rm=TRUE),
             Maximum = max( Turbidity_ntu, na.rm=TRUE) ) %>%
  kable( caption = "Water turbidity from sensors at the Rice Rivers Center, in the first quarter of 2014.") %>%
  kable_classic_2( full_width=FALSE) %>%
  add_header_above( c(" ", "Turbidity"=2))
  
```







