---
title: "Raster Data Homework"
format: html
author: "Rebecca Hoover"
title-block-style: default
title-block-banner: images/beetle-skin.png
date: "10/31/2024"
abstract: "Rasters are continuous spatial representations of geospatial data. In this exercise, we practiced working with raster data either directly or using dplyr and standard data manipulation techniques to load, manipulate, and visualize the data. In doing so, we used beetle data collected from the cape region of Mexico"
smooth-scroll: true
code-tools: true
toc: false
editor_options: 
  chunk_output_type: console
chunk_output_type: console
---

## Data and Loaded Packages

```{r, message=FALSE, warning=FALSE}
#Load in Packages
library(tidyverse)
library(raster)
library(sf)
library(ggrepel)

raster_url <- "https://github.com/DyerlabTeaching/Raster-Data/raw/main/data/alt_22.tif"

beetle_url <- "https://raw.githubusercontent.com/DyerlabTeaching/Raster-Data/main/data/AraptusDispersalBias.csv"
```

## Running Code

1. Load the raster and point data in and crop to an appropriate size to display the locations of the sampling plots and make a label for each site.

```{r, message=FALSE, warning=FALSE}

elev <- raster( raster_url )

read_csv( beetle_url )  %>% 
  st_as_sf( coords = c("Longitude","Latitude"), 
            crs = 4326 ) -> beetles

plot( elev, xlab="Longitude" )

beetles %>%
  st_union() %>%
  st_buffer( dist = 1 ) %>%
  st_bbox() 

baja_extent <- extent( c(-115.29354,-108,22,30))

baja_elev <- crop( elev, baja_extent )
baja_elev

plot( baja_elev) 

plot( baja_elev, xlab="Longitude", ylab="Latitude")
text( beetles, add=TRUE, pch=16, col="red" )

baja_elev %>%
  rasterToPoints() %>%
  as.data.frame() %>% 
  ggplot() +
  geom_raster( aes(x,y,fill=alt_22) ) +
  geom_sf_label( aes(label=Site), 
                 data = beetles) +
  coord_equal()
```
2. Use the click() function to crop the raster and filter the sites to include only the region that is on the peninsula of Baja California. Plot the raster and sites as labels.
```{r}

cape_bound <- c(-116, -109, 22, 30)
cape_ex <- extent(cape_bound)

# Crop the raster to the Cape region
cropped_rast <- crop(baja_elev, cape_ex)

# Filter for beetle sites south of Site #75
cape_pts <- beetles %>% filter(Site >= 75)

# Ensure 'cape_points' is an sf object and contains coordinates
if (!inherits(cape_pts, "sf")) {
  cape_pts <- st_as_sf(cape_pts, coords = c("Longitude", "Latitude"), crs = 4326)
}

# Plot the raster for the Cape region
plot(cropped_rast, xlab = "Longitude", ylab = "Latitude", main = "Cape Region")

max_mfratio <- max(cape_pts$MFRatio, na.rm = TRUE)
scaled_cex <- cape_pts$MFRatio / max_mfratio

scaled_cex[is.na(scaled_cex)] <- 1

plot(cape_pts, add = TRUE, col = "red", pch = 16, cex = scaled_cex)

text(st_coordinates(cape_pts), labels = cape_pts$Site, pos = 3, cex = 0.7)
```
3. The peninsula of Baja California is divided into the States of Baja California Norte and Baja California Sur. The border between these states is at 28° Latitude. Divide the sample locations into groups based upon which state they are located and plot the average sex ratio of the sample sites partitioned by each site.
```{r, message=FALSE, warning=FALSE}
beetles <- beetles %>%
  mutate(
    Latitude = st_coordinates(beetles)[, 2],
    State = case_when(
      Latitude > 28 ~ "Baja California Norte", 
      TRUE ~ "Baja California Sur"
    )
  )

# Calculate average M/F ratio by state
state_avg_sex_ratio <- beetles %>%
  group_by(State) %>%
  summarize(AverageMFRatio = mean(MFRatio, na.rm = TRUE))


# Bar ggplot for average M/F ratio by state
        
ggplot(state_avg_sex_ratio, aes(x = State, 
                                y = AverageMFRatio, 
                                fill = State)) +
  geom_bar(stat = "identity", 
           color = "black", width = 0.7) + 
  labs(title = "Average M/F Ratio by State", 
       y = "Average M/F Ratio", 
       x = "State") +
  scale_fill_manual(values = c("red", "blue")) +
  theme(legend.position = "none")
```


4. Is there a relationship between the observed sex ratio and the elevation at that site? Plot these, and in the text, perform a correlation test (cor.test()) and report the values as if you were writing them up in a results section.

```{r, message=FALSE, warning=FALSE}
# Extract elevation values at each sampling location
elevation_v <- raster::extract(cropped_rast, beetles)
beetles <- beetles %>% mutate(Elevation = elevation_v)

# Correlation test between MFRatio and Elevation
cor_test <- cor.test(beetles$MFRatio, beetles$Elevation, method = "pearson")

# Scatter plot 

ggplot(beetles, aes(x = Elevation, y = MFRatio)) +
  geom_point() +  
  geom_smooth(method = "lm", color = "red", size = 1, se = FALSE) +  
  labs(
    title = "M/F Ratio vs. Elevation",
    x = "Elevation (m)",
    y = "M/F Ratio",
    subtitle = paste("Correlation: r =", round(cor_test$estimate, 2), ", p =", format.pval(cor_test$p.value))
  ) +
  theme() 

# Print correlation test results
cat("The correlation between M/F ratio and elevation was r =", round(cor_test$estimate, 2), ", p =", cor_test$p.value, "\n")


```
