---
title: "RDS Assignment"
format: html
author: "Rebecca Hoover"
title-block-style: default
title-block-banner: true
date: "11/15/2024"
abstract: "This week, the class covered downloading and calibrating Ecotone RDS groundwater monitoring wells, noting their out-of-date in favor of Onset Hobo pressure transducers. The upcoming focus will be transforming and analyzing raw data from Meazura to visually display groundwater data from the Kimages Creek wetlands restoration project. This is what we are doing below."
smooth-scroll: true
code-tools: true
toc: false
editor_options: 
  chunk_output_type: console
chunk_output_type: console
---

# Load Packages and Clean Data

```{r, message=FALSE, warning=FALSE}

# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(RColorBrewer)
library(readr)
library(lubridate)


# Define the file paths
w1_file_path <- "/Users/rebeccahoover/Downloads/Data Literacy/EDL-ENVS543/Hoover_Environmental-Data-Literacy_ENVS-543/000013D49BD9_20240228_1149.txt"

w3_file_path <- "/Users/rebeccahoover/Downloads/Data Literacy/EDL-ENVS543/Hoover_Environmental-Data-Literacy_ENVS-543/00000A3C04D2_20240225_1231.txt"

w4_file_path <- "/Users/rebeccahoover/Downloads/Data Literacy/EDL-ENVS543/Hoover_Environmental-Data-Literacy_ENVS-543/000013151FF7_20240225_1158.txt"

# Function to clean up the file
clean_file <- function(file_path) {
  # Read the file as lines
  lines <- read_lines(file_path)
  
  # Since metadata spans the first 4 lines
  start_line <- 5  #Extracts after metadata
  
  # Extract only the CSV data
  csv_data <- lines[start_line:length(lines)]
  
  # Join the lines back into a single string with newline characters
  csv_text <- paste(csv_data, collapse = "\n")
  
  # Write the cleaned CSV data to a new file
  cleaned_file_path <- paste0(sub(".txt$", "", file_path), "_cleaned.csv")
  write_lines(csv_text, cleaned_file_path)
  
  return(cleaned_file_path)
}

# Clean each file
cleaned_well_1 <- clean_file(w1_file_path)
cleaned_well_3 <- clean_file(w3_file_path)
cleaned_well_4 <- clean_file(w4_file_path)

```

# Wrangle Data

```{r, message=FALSE, warning=FALSE}

# Read in
well_1_data <- read_csv(cleaned_well_1)
well_3_data <- read_csv(cleaned_well_3)
well_4_data <- read_csv(cleaned_well_4)

#Rename Columns
well_1_data <- well_1_data %>%
  rename(Date = `____Date___`, Level = `_Level_`)

well_3_data <- well_3_data %>%
  rename(Date = `____Date___`, Level = `_Level_`)

well_4_data <- well_4_data %>%
  rename(Date = `____Date___`, Level = `_Level_`)

# Convert Date 
well_1_data <- well_1_data %>%
  mutate(Date = mdy(Date), Level = as.numeric(Level))

well_3_data <- well_3_data %>%
  mutate(Date = mdy(Date), Level = as.numeric(Level))

well_4_data <- well_4_data %>%
  mutate(Date = mdy(Date), Level = as.numeric(Level))

# Add a "Well" column 
well_1_data$Well <- "Well 1"
well_3_data$Well <- "Well 3"
well_4_data$Well <- "Well 4"

# Add 6 inches to the ( Well 1 ) Level (Depth) column
well_1_data <- well_1_data %>%
  mutate(Level = Level + 6)

# Combine the three well datasets into one
combined_data <- bind_rows(well_1_data, well_3_data, well_4_data)

#Filter for Year
combined_data_2023 <- combined_data %>%
  filter(year(Date) == 2023)

filtered_data <- combined_data_2023 %>%
  filter(Date >= as.Date("2023-03-28") & Date <= as.Date("2023-11-06"))

```

# Graphics

```{r}
# Create a line plot 
ggplot(filtered_data, aes(x = Date, 
                          y = Level, 
                          color = Well, 
                          group = Well)) +
  geom_line() + 
  labs(title = "Depth of Ground Water Table for Wells: W-1, W-3 and W-4.\nDuring the 2023 Growing Season: March 28th - November 8th",
       x = "Date",
       y = "Depth to Groundwater Table (inches)", 
       color = NULL) +  
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5, family = "Times New Roman"),  
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, family = "Times New Roman"),
        axis.text.y = element_text(family = "Times New Roman"),
        axis.title.x = element_text(family = "Times New Roman"),
        axis.title.y = element_text(family = "Times New Roman"),
        legend.text = element_text(family = "Times New Roman")) +    
  scale_color_brewer(palette = "Set2") +     #color blind friendly
  geom_hline(yintercept = 0, linetype = "dotted", color = "black")+
  geom_hline(yintercept = -12, color = "black")+
  geom_vline(xintercept = as.Date("2023-03-28"), color = "black") +
  geom_text(aes(x = as.Date("2023-03-28"), y = min(Level) - 5,
                label = "Start of Growing Season (2023-03-28)"), 
            angle = 90, vjust = - 1, hjust = -0.19, color = "black", family = "Times New Roman") + 
  scale_x_date(breaks = "1 month",  
               labels = scales::date_format("%b %d, %Y"))



```

# Discussion

To meet the ACOE hydrology success criteria, the water levels at these sites must remain at or above 1 foot from the soil surface for at least 12.5% of the growing season. From the beginning of the growing season through mid-May, all three wells consistently meet this criterion. However, upon reviewing the raw data, a decline in water levels to 1 foot below surface level is observed starting on May 23, 2023, for Well 1.

The growing season spans 225 consecutive days, with 56 days occurring between the start of the season and May 23rd. For this 56-day period, Well 1 maintains water levels above 1 foot from the surface, meeting the 12.5% criterion for that portion of the season. Given that the total growing season lasts 225 days, Well 1 meets the ACOE hydrology success criterion for at least 12.5% of the growing season up to May 23, 2023.

Therefore, all three wells satisfy the hydrology requirements set forth by the ACOE.
